---
title: "440 CS3 IR2"
author: "Jake Epstein, Man-Lin Hsiao, Sahil Patel, Daniel Spotiswood, & Michael Tan"
date: "10/16/2019"
output: html_document
---

```{r, echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(lme4)
library(sjPlot)
library(brms)

select = dplyr::select
```

```{r, echo=FALSE}
#read in mortality
mortality = read.csv("nc_infant_mortality.csv") %>%
  mutate(COUNTY = str_remove_all(COUNTY, " "))

 
mortality$COUNTY[which(mortality$COUNTY == "McDOWELL")] = "MCDOWELL"
mortality$COUNTY[which(mortality$COUNTY == "TYRELL")] = "TYRRELL"

demos = read.csv("Data/County_Demo_Data.csv")
demos.clean = demos %>%
  mutate(COUNTY = toupper(COUNTY) %>% str_remove_all(" "))

mortality = merge(mortality , demos.clean, by = "COUNTY")
```

```{r, echo=FALSE}
# replicate 2015 SCHS table exactly
SCHS_Table_2015 = mortality %>%
  mutate(race = relevel(race, "White")) %>%
  filter(YOB == 2015) %>%
  select(mortality, race, COUNTY, deaths) %>%
  mutate(smallSample = deaths < 10) %>%
  select(-deaths, -smallSample) %>%
  spread(race, mortality)
```

### EDA
```{r, echo=FALSE}
#Drop Clay outlier for plots
eda=mortality%>%
  filter(mortality<1000)
```

XXXtableau screenshot

XXX


```{r, echo=FALSE}
#Median income
cor1=cor(eda$Median_Household_Income,eda$mortality_rate)
ggplot(eda)+
  aes(x=Median_Household_Income,y=mortality_rate)+
  geom_point()+
  geom_smooth(method=lm)+
  theme_minimal()+
  labs(title="Median Household Income v. Mortality Rate",y="Mortality Rate",x="Median Household Income")
```
As illustrated by the scatter plot above, there is distinct negative relationship between the median income of a county and the mortality rate in that county. Specifically, the two variables have a negative correlation of `r cor1`. This implies that counties with higher median income tend to have lower mortatility rates.

```{r, echo=FALSE}
#Low birth weight
cor2=cor(eda$pctLowBirthWeight,eda$mortality)
ggplot(eda)+
  aes(x=pctLowBirthWeight,y=mortality_rate)+
  geom_point()+
  geom_smooth(method=lm)+
  theme_minimal()+
  labs(title="Percent Low Birth Weight v. Mortality Rate",y="Mortality Rate",x="Percent Low Birth Weight")
```
As illustrated by the scatter plot above, there is distinct negative relationship between the percent of births that had a low birth weight in a county and the mortality rate in that county. Specifically, the two variables have a positive correlation of `r cor2`. This implies that counties with a percent of births that had a low birth weight tend to have higher mortatility rates.

```{r, echo=FALSE}
#Percent Smoke
cor3=cor(eda$pctSmokers,eda$mortality_rate)
ggplot(eda)+
  aes(x=pctSmokers,y=mortality_rate)+
  geom_point()+
  geom_smooth(method=lm)+
  theme_minimal()+
  labs(title="Percent Smokers v. Mortality Rate",y="Mortality Rate",x="Percent Smokers")
```
As illustrated by the scatter plot above, there is distinct negative relationship between the proportion of a county that smokes and the mortality rate in that county. Specifically, the two variables have a negative correlation of `r cor3`. This implies that counties with more smokers tend to have lower mortatility rates.

### Model Selection

XXX

```{r, echo=FALSE}
mortality = mortality %>%
  mutate(mortality_rate = mortality / 1000, race = relevel(race, "White")) %>%
  mutate(Median_Household_Income = Median_Household_Income / 1000)

model = glmer(data = mortality, weights = births, formula = mortality_rate ~ (1|COUNTY) + (1|race) + Median_Household_Income + pctLowBirthWeight + pctSmokers, family = binomial())

summary(model)
plot_model(model)

plot_model(model, type = "re") 
plot_model(model, type = "est")
```
These charts show the differing intercepts for the differing counties and races. 

##Model Interpretations
The model revealed the following statistically significant effects:

For a given county, race, and percentage of births in a county that have low birth weight, a one thousand dollar increase in median household income leads to 0.975x the probability of mortality (changed by a multiplicative factor of e^-0.025122  = 0.975).

The explanation of this seems intuitive. It could be the case that becoming wealthier leads to being able to afford better obstetrics (birthing and pregnancy) services, which would lead to a decreased probability of a baby dying.

For a given county, race, and median household income in a county, a one percent increase in the percentage of births that have low birth weight leads to 0.954x the probability of mortality (changed by a multiplicative factor of e^-0.046546  = 0.954).

The explanation of this seems counterintuitive. It could be the case that families that know they will have a baby with low birth weight (mothers that know they will give birth early) will prepare for such a situation by picking a higher-quality obstetrics service, which would lead to a decreased probability of a baby dying.

The county level variability is 0.01728 and the race level variability is 0.11702.

### Comparison
```{r, echo=FALSE, warning=FALSE}
mortality_fitted = mortality
mortality_fitted$fitted.values = fitted(model)


our_table = mortality_fitted %>%
  filter(YOB == 2015) %>%
  select(fitted.values, race, COUNTY) %>%
  mutate(predicted_mortality = 1000*fitted.values) %>%
  select(-fitted.values) %>%
  spread(race, predicted_mortality)


our_table_as_list = mortality_fitted %>%
  select(COUNTY, race, fitted.values) %>%
  mutate(predicted_mortality = 1000*fitted.values) %>%
  select(-fitted.values)

schs_table_as_list = SCHS_Table_2015 %>%
  gather(race, mortality, 2:length(SCHS_Table_2015)) %>%
  na.omit()

table_comparisons = merge(our_table, SCHS_Table_2015, by = "COUNTY")

comparisons = merge(our_table_as_list, schs_table_as_list, by = c("race", "COUNTY"))

comparisons = merge(comparisons, mortality %>% filter(YOB == 2015))

comparisons = comparisons %>%
  mutate(difference = mortality - predicted_mortality)

write.csv(comparisons, "comparisons.csv")

ggplot(comparisons, aes(x = predicted_mortality, y = mortality)) +
  geom_point() +
  ylim(c(0, 100)) +
  geom_abline(slope = 1, intercept = 0) +
  xlab ("Predicted Values") +
  ylab ("Reported Values") +
  theme_minimal()

ggplot(comparisons, aes(x = births, y = difference)) +
  geom_point() +
  ylim(c(-10, 100)) +
  xlab("Number of Births") +
  ylab("Difference in Predicted vs Reported Mortalities") +
  theme_minimal()
```
We see that our predictions do a strong job of estimating the reported values, and are centered around a slope of 1, so there is no clear bias The large residual points are mostly indicative of reported values that draw on a small quantity of data points. We must note that there are some points that lie outside of this graph, but again are mostly caused by a lack of data. This is demonstrated in the third chart that shows that as the number of births increase our predictions converge with the sample means.