---
title: "440cs3_ir2"
author: "Jake Epstein"
date: "10/16/2019"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(lme4)
library(sjPlot)
library(brms)

select = dplyr::select
```

```{r}
#read in mortality
mortality = read.csv("nc_infant_mortality.csv") %>%
  mutate(COUNTY = str_remove_all(COUNTY, " "))

 
mortality$COUNTY[which(mortality$COUNTY == "McDOWELL")] = "MCDOWELL"
mortality$COUNTY[which(mortality$COUNTY == "TYRELL")] = "TYRRELL"

demos = read.csv("Data/County_Demo_Data.csv")
demos.clean = demos %>%
  mutate(COUNTY = toupper(COUNTY) %>% str_remove_all(" "))

mortality = merge(mortality , demos.clean, by = "COUNTY")
```


```{r}
# replicate 2015 SCHS table exactly

SCHS_Table_2015 = mortality %>%
  mutate(race = relevel(race, "White")) %>%
  filter(YOB == 2015) %>%
  select(mortality, race, COUNTY, deaths) %>%
  mutate(smallSample = deaths < 10) %>%
  select(-deaths, -smallSample) %>%
  spread(race, mortality)


```


```{r}
mortality = mortality %>%
  mutate(mortality_rate = mortality / 1000, race = relevel(race, "White")) %>%
  mutate(Median_Household_Income = Median_Household_Income / 1000)

model = glmer(data = mortality, weights = births, formula = mortality_rate ~ (1|COUNTY) + (1|race) + Median_Household_Income + pctLowBirthWeight, family = binomial())

summary(model)
plot_model(model)

plot_model(model, type = "re") 
plot_model(model, type = "est") 


```
These charts show the differing intercepts for the differing counties and races. 



### Compare our fitted values to the predictions from the schs table
```{r}
mortality_fitted = mortality
mortality_fitted$fitted.values = fitted(model)


our_table = mortality_fitted %>%
  filter(YOB == 2015) %>%
  select(fitted.values, race, COUNTY) %>%
  mutate(predicted_mortality = 1000*fitted.values) %>%
  select(-fitted.values) %>%
  spread(race, predicted_mortality)


our_table_as_list = mortality_fitted %>%
  select(COUNTY, race, fitted.values) %>%
  mutate(predicted_mortality = 1000*fitted.values) %>%
  select(-fitted.values)

schs_table_as_list = SCHS_Table_2015 %>%
  gather(race, mortality, 2:length(SCHS_Table_2015)) %>%
  na.omit()

table_comparisons = merge(our_table, SCHS_Table_2015, by = "COUNTY")

comparisons = merge(our_table_as_list, schs_table_as_list, by = c("race", "COUNTY"))

comparisons = merge(comparisons, mortality %>% filter(YOB == 2015))

comparisons = comparisons %>%
  mutate(difference = mortality - predicted_mortality)

write.csv(comparisons, "comparisons.csv")

ggplot(comparisons, aes(x = predicted_mortality, y = mortality)) +
  geom_point() +
  ylim(c(0, 100)) +
  geom_abline(slope = 1, intercept = 0) +
  xlab ("Predicted Values") +
  ylab ("Reported Values") +
  theme_minimal()

ggplot(comparisons, aes(x = births, y = difference)) +
  geom_point() +
  ylim(c(-10, 100)) +
  xlab("Number of Births") +
  ylab("Difference in Predicted vs Reported Mortalities") +
  theme_minimal()
```
We see that our predictions do a strong job of estimating the reported values, and are centered around a slope of 1, so there is no clear bias The large residual points are mostly indicative of reported values that draw on a small quantity of data points. We must note that there are some points that lie outside of this graph, but again are mostly caused by a lack of data. This is demonstrated in the third chart that shows that as the number of births increase our predictions converge with the sample means.






